---
title: "p8105_hw2_yl5829"
output: html_document
date: "2024-10-01"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readxl)
```

## R Markdown

#Problem 1

```{r}
#load the data in relative path
nyc <- read_csv("data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv", na = c(".", "NA", ""))
head(nyc)
```


```{r}
#clean the data
nyc_clean <- nyc %>%
  janitor::clean_names() %>%
  select(line, station_name, station_latitude, station_longitude, 
         route1, route2, route3, route4, route5, route6, route7, route8, route9, route10, route11, 
         entry, vending, entrance_type, ada) %>%
  mutate(entry = ifelse(entry == "YES", TRUE, FALSE))

#look at the cleaned data
head(nyc_clean)
```

```{r}
#check for column names
colnames(nyc)
#check for number of column
ncol(nyc)
#check for dimension of cleaned data
dim(nyc_clean)
```

The dataset contains 32 variables and it mainly focuses on the subway entrances and exits in the NYC transit system. The variables included are line name, station name, geographic information (latitude and longitude), various routes served by the station (up to 11 routes), entry type, vending machine availability, some street, entrance type and geographic information (latitude, longitude, and location), and ADA compliance. After loading the data I use janitor::clean_names() function to convert all the variable to lower case and replaced spaces with underscores  to better performe next step. I next select relevant variables and transformed the entry column, which initially contained "YES" and "NO" values, into a logical variable (TRUE for "YES" and FALSE for "NO"). I also handled the missing values by specify any missing value to be NA in the data.
After clean there are 1868 rows and 19 columns in total. The data is tidy as each variable is stored in its own column, each observation is in its own row, and the dataset is organized for further analysis.

```{r}
#the number of distinct stations
distinct_stations <- nyc_clean %>%
  distinct(station_name, line) %>%
  nrow()

#see the output
distinct_stations
```
There are 465 distinct stations.

```{r}
#the number of ADA compliant stations
ada_compliant <- nyc_clean %>%
  filter(ada == "TRUE") %>%
  distinct(station_name, line) %>%
  nrow()

#see the output
ada_compliant
```
There are 84 ADA compliant.

```{r}
#proportion of station entrances / exits without vending allow entrance
proportion <- nyc_clean %>%
  filter(vending == "NO" & entry == TRUE) %>%
  nrow() / nrow(nyc_clean)

# Output the proportion
proportion
```
There is 0.0369 station entrances / exits without vending allow entrance

```{r}
#convert all route columns to character type
nyc_clean <- nyc_clean %>%
  mutate(across(starts_with("route"), as.character))

#reshape the data
nyc_routes <- nyc_clean %>%
  pivot_longer(cols = starts_with("route"), names_to = "route_number", values_to = "route_name") %>%
  filter(!is.na(route_name))

#look at the reshaped data
head(nyc_routes)
```

```{r}
#count distinct stations that serve the A train
a_train <- nyc_routes %>%
  filter(route_name == "A") %>%
  distinct(station_name, line) %>%
  nrow()

#number of distinct stations that serve the A train
a_train

#count ADA compliant stations that serve the A train
a_ada <- nyc_routes %>%
  filter(route_name == "A", ada == TRUE) %>%
  distinct(station_name, line) %>%
  nrow()

#number of ADA compliant stations that serve the A train
a_ada
```
There are 60 distinct stations serve the A train and of them 17 are ADA compliant

#Problem 2

```{r}
#load the mr trash wheel data in relative path
mr_wheel <- read_excel("data/202409 Trash Wheel Collection Data.xlsx", sheet = "Mr. Trash Wheel", na = c(".", "NA", ""))
head(mr_wheel)
```

```{r}
mr_wheel_clean <- mr_wheel %>%
  #make more reasonable names
  rename(
    dumpster = `Dumpster`,
    month = `Month`,  
    year = `Year`,    
    date = `Date`,
    weight_tons = `Weight (tons)`,
    volume_cubic_yards = `Volume (cubic yards)`,
    plastic_bottles = `Plastic Bottles`,
    polystyrene = `Polystyrene`,
    cigarette_butts = `Cigarette Butts`,
    glass_bottles = `Glass Bottles`,
    sports_balls = `Sports Balls`,
    homes_powered = `Homes Powered*`
  ) %>%
  mutate(year = as.character(year))%>%
  #remove rows that don't contain dumpster-specific data
  filter(!is.na(dumpster)) %>%
  #round the number of sports balls to the nearest integer and convert to integer
  mutate(sports_balls = as.integer(round(sports_balls, 0)))

#look at the cleaned data
head(mr_wheel_clean)
```

```{r}
#professor Trash Wheel data
prof_wheel <- read_excel("data/202409 Trash Wheel Collection Data.xlsx", sheet = "Professor Trash Wheel")

head(prof_wheel)
```

```{r}
#clean Professor Trash Wheel data
prof_wheel_clean <- prof_wheel %>%
  rename(
    dumpster = `Dumpster`,
    month = `Month`,
    year = `Year`,
    date = `Date`,
    weight_tons = `Weight (tons)`,
    volume_cubic_yards = `Volume (cubic yards)`,
    plastic_bottles = `Plastic Bottles`,
    polystyrene = `Polystyrene`,
    cigarette_butts = `Cigarette Butts`,
    glass_bottles = `Glass Bottles`,
    plastic_bags = `Plastic Bags`,
    wrappers = `Wrappers`,
    homes_powered = `Homes Powered*`
  ) %>%
  mutate(year = as.character(year)) %>%
  filter(!is.na(dumpster))

head(prof_wheel_clean)
```
```{r}
#add a new column to identify the data source
prof_wheel_clean <- prof_wheel_clean %>%
  mutate(source = "Professor Trash Wheel")
prof_wheel_clean
```

```{r}
#Gwynnda Trash Wheel data
gwynnda_wheel <- read_excel("data/202409 Trash Wheel Collection Data.xlsx", sheet = "Gwynnda Trash Wheel")
head(gwynnda_wheel)
```

```{r}
#clean Gwynnda Trash Wheel data
gwynnda_wheel_clean <- gwynnda_wheel %>%
    rename(
    dumpster = `Dumpster`,
    month = `Month`,
    year = `Year`,
    date = `Date`,
    weight_tons = `Weight (tons)`,
    volume_cubic_yards = `Volume (cubic yards)`,
    plastic_bottles = `Plastic Bottles`,
    polystyrene = `Polystyrene`,
    cigarette_butts = `Cigarette Butts`,
    plastic_bags = `Plastic Bags`,
    wrappers = `Wrappers`,
    homes_powered = `Homes Powered*`
  ) %>%
  mutate(year = as.character(year))%>%
  #remove rows that don't contain dumpster-specific data
  filter(!is.na(dumpster))

#loot at the cleaned data
head(gwynnda_wheel_clean)
```
```{r}
#add a new column to identify the data source (Gwynnda Trash Wheel)
gwynnda_wheel_clean <- gwynnda_wheel_clean %>% mutate(source = "Gwynnda Trash Wheel")
head(gwynnda_wheel_clean)
```

```{r}
#add a new column to identify the data source (Mr. Trash Wheel)
mr_wheel_clean <- mr_wheel_clean %>%
  mutate(source = "Mr. Trash Wheel")
head(mr_wheel_clean)
```

```{r}
#combine all datasets into one tidy dataset
combined_wheel <- bind_rows(mr_wheel_clean, prof_wheel_clean, gwynnda_wheel_clean) %>% 
  janitor::clean_names()

#look at the combined dataset
head(combined_wheel)
```

The dataset combines data from three datasets: Mr. Trash Wheel, Professor Trash Wheel, and Gwynnda Trash Wheel. There are a total of `r nrow(combined_wheel)` observations. Each observation represents a specific dumpster, with key variables such as weight_tons, plastic_bottles, polystyrene, and cigarette_butts. The total weight of trash collected by Professor Trash Wheel is `r sum(subset(combined_wheel, source == "Professor Trash Wheel")$weight_tons, na.rm = TRUE)` tons. The total number of cigarette butts collected by Gwynnda Trash Wheel in June 2022 is `r sum(subset(combined_wheel, source == "Gwynnda Trash Wheel" & format(date, "%Y-%m") == "2022-06")$cigarette_butts, na.rm = TRUE)`.

#Problem 3

```{r}
#import the dataset
bakers <- read_csv("data/bakers.csv")
bakes <- read_csv("data/bakes.csv")
results <- read_csv("data/results.csv")

#look through dataset
head(bakers)
head(bakes)
head(results)
```
```{r}
#make results dataset more readable and reasonable
results <- results %>%
  rename(series = 1, episode = 2, baker = 3, technical = 4, result = 5) %>% 
  slice(-1) %>%
  filter(!is.na(series))
head(results)
```

```{r}
#extract first names from Baker Name in `bakers`, and make data more reasonable to read
bakers_clean <- bakers %>%
  separate(`Baker Name`, into = c("First Name", "Last Name"), sep = " ", extra = "merge", fill = "right") %>%
  mutate(baker = tolower(`First Name`)) %>%
  rename(series = `Series`,
         baker_age = `Baker Age`,
         baker_occupation = `Baker Occupation`,
         hometown = `Hometown`) %>% 
  select(baker, series, baker_age, baker_occupation, hometown)

print(bakers_clean)
```

```{r}
#clean `bakes` dataset
bakes_clean <- bakes %>%
  mutate(Baker = tolower(Baker),       
         Series = tolower(Series),     
         Episode = as.character(Episode)) %>%
  rename(series = `Series`, 
         episode =`Episode`,
         baker = `Baker`,
         signature_bake = `Signature Bake`,
         show_stopper = `Show Stopper`)

print(bakes_clean)
```

```{r}
#clean `results` dataset
results_clean <- results %>%
  mutate(baker = tolower(baker),     
         series = tolower(series),    
         episode = as.character(episode))

print(results_clean)
```

```{r}
#check if unmatched bakers in `bakes`
unmatched_bakes <- bakes_clean %>%
  anti_join(bakers_clean, by = c("baker", "series"))

#check if unmatched bakers for `results`
unmatched_results <- results_clean %>%
  anti_join(bakers_clean, by = c("baker", "series"))

print(unmatched_bakes)
print(unmatched_results)
```
By checking it seems there are 8 unmatched results. However, after observation they are all from same person just that the inconsistent name conventions, jo stands for joannes (they show up in smae series and episode and has similar name short cut so we conlcude that).

```{r}
#remove any extra quote mark in the `baker` 
bakes_clean <- bakes_clean %>%
  mutate(baker = gsub('"', '', baker))  # Remove double quotes

#replace jo with joanne in bakers_clean
bakers_clean <- bakers_clean %>%
  mutate(baker = ifelse(baker == "jo", "joanne", baker))

#check if the replacement work
print(bakers_clean %>% filter(baker == "joanne"))

```


```{r}
#check if unmatched bakers in `bakes`
unmatched_bakes <- bakes_clean %>%
  anti_join(bakers_clean, by = c("baker", "series"))

#check if unmatched bakers for `results`
unmatched_results <- results_clean %>%
  anti_join(bakers_clean, by = c("baker", "series"))

print(unmatched_bakes)
print(unmatched_results)
```
the result looks good here has no unmatched data, the datasets are complete and correct.


```{r}
#merge bakers and results on baker and series
bakers_results <- full_join(bakers_clean, results_clean, by = c("baker", "series"))

#merge bakers with bakes on baker, series, and episode
final_dataset <- full_join(bakers_results, bakes_clean, by = c("baker", "series", "episode"))

#see the final dataset
print(final_dataset)
```



```{r}
write_csv(final_dataset, "data/final_baking_dataset.csv")
```

For data cleaning process, after loading the data i found the results dataset look unreasonable as the column name were represented by numbers and first row is name of the column, so I remove the first row and change the name of the column. Next, I’ve noticed the baker in bakers dataset is formed as First Name plus Last Name, which is completely inconsistent to other dataset, so I change that by create a new column that only contains baker’s first name. and I rename all the column names to lower letters or with _ between words and I did same procedure to other two datasets. I next checked whether the datasets is matched or not and found 8 unmatched row because of the inconsistent name conventions, and I fix it by remove the conventions part and make all name consistent. Then I retest for matches and found all rows are matched. Lastly I merge all the dataset and write a csv final document. The choice I make is that I opted to correct mismatched names (like jo vs. joanne) rather than removing or excluding them, to preserve the completeness of the data. Another one is that instead of using inner join I use full join as inner join will dismiss series 9 and 10.
The final dataset includes all relevant details from the bakers, bakes, and results datasets. Each row contains the baker, series, episode, baker_age, baker_occupation, hometown, their signature bake, show stopper, technical score, and final result for that episode.

```{r}
#filter the final dataset for Seasons 5 through 10 and extract star bakers or winners
star_bakers <- final_dataset %>%
  filter(series %in% c(5, 6, 7, 8, 9, 10)) %>%
  filter(result == "STAR BAKER" | result == "WINNER") %>%
  select(series, episode, baker, result) %>%
  arrange(series, episode)

#see the star baker or winner table
print(star_bakers)
```


